\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-28 paper/algorithm.tex
paper: pass over algorithms

\section{Algorithm}

\todobox{this jumps around too much}{
Our optimizer searches (\Cref{sec:planning}) a restricted space (\Cref{sec:search-space}) of query plans containing imputation operators (\Cref{sec:operators}) for an optimal plan given an estimate of the runtime of the query, information lost through imputation operations, and a parameter trading off the two.

This plan is subject to two constraints (\Cref{sec:placement}).
First, it must not emit any tuples which contain null values, regardless of the state of the base tables.
Second, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) must never observe a null value in any attribute that they operate on directly.
Our planning algorithm is agnostic to the type of imputation used (\Cref{sec:imputation}), as the properties of the imputation are summarized by the cost model (\Cref{sec:cost-model}).}

@@

\subsection{Cost model}
\label{sec:cost-model}
The cost of a plan $q$ is expressed as a tuple $\langle \textsc{Loss}(q), \textsc{Time}(q) \rangle$.
$\textsc{Time}(q) \in (0, \infty)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates. $\textsc{Loss}(q) \in [0,1]$ is an estimate of the amount of information lost by the imputation procedures used.

@@

This set $P$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all possible trade-offs of performance and cost for the current query.

@@

The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ relies on having an accurate cardinality estimate of a query plan $q$ and its sub-plans. 
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
For simplicity, each of the logical nodes in a query plan points to a set of histograms.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
\Cref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.

For each sub-plan, we keep track of the estimated number of
null values in a column  ($\textsc{Missing}(c, Q)$) by using the histograms associated with
that plan. We use this estimate to compute $\text{Loss}(Q)$, a measure averaged over imputation operations, as follows.

@@

To compute $\text{Time}(Q)$, an additive measure of runtime cost, we retain a simple set of heuristics used by a standard 
database.

@@

In general, evaluating the time complexity for different \textit{Impute} operators is
complicated, as the properties of
different algorithms can vary significantly, the underlying I/O cost
is not easily extracted (especially in the case that the entire stream of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
not even consider CPU cost explicitly). For example, \cite{martin1995time} find that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). Indeed, the \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree. In
practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
plans).

@@

Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a null value, there are cases where an imputation operator is required. To track these cases, each query plan $q$ is associated with a set of dirty attributes $\textsc{Dirty}(q)$. An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain null. We compute a dirty set for each table using the table statistics, which track the number of null values in each column. The dirty set for a query plan can be computed recursively as follows:

@@

Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
At a high level, a plan cache is a mapping from a set of tables to a set of dominating plans that perform the required joins and filters over these tables.
We store multiple plans rather than a single plan because each plan has a distinct set of dirty attributes and a choice over the plan frontier
can only be made once the set is final. Details are shown in~\Cref{fig:semantics-plan-cache}.

\begin{figure}
  \begin{align*}
    \llbracket Q[T] \rrbracket &= \begin{cases}
      P & (T, P) \in Q \\
      \emptyset & \text{otherwise} \\
    \end{cases} \\
    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned} Q \cup \{(T,\ P') \} \end{aligned}\\
                                            \text{ where } \\
                                           P' &= \left\{q ~|~ \begin{aligned}
                                                                     q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\
                                                                     \textsc{Dirty}(q) = \textsc{Dirty}(p),\\
                                                                     \forall q' \in \llbracket Q[T] \rrbracket.  \doms{q} {q'} 
                                                                   \end{aligned}
                                           \right\}
  \end{align*}
  \caption{Semantics of a plan cache $Q$}
  \label{fig:semantics-plan-cache}
\end{figure}

@@

The join optimization is based on the algorithm used in System R\cite{blasgen1981system},
but rather than select the best plan at every point, we use our plan cache, which maintains the Pareto frontier of possible plans. The output of the join optimizer is the set of plans which satisfy all join predicates and lie along the frontier.

@@

The flexibility of this non-parametric imputation approach aligns well with the spirit of
query optimization, which aims to reconcile performance with a declarative interface to data
manipulation. By removing concerns for imputation, users executing queries in \ProjectName{}
do not need to consider the potential nature of missing data.

@@

\subsection{Optimality}
We prove the optimality of the final plan selected by showing that our cost-model and algorithm
construct a complete Pareto frontier.

\begin{definition}
$I: \mathcal{P}(D) \rightarrow \mathcal{P}(\{ \delta, \mu \})$, where
$D$ is the set of dirty attributes for a relation, is an abstract imputation operator mapping a set of dirty
attributes to a set of imputation operations.

@@

\begin{theorem}\label{theorem:no-bad-sit}
By computing dirty-set specific Pareto frontiers, it is never the case that we discard a plan $p_1$ s.t. $\doms{p_2}{p_1}$, but $\exists p_1' \in I(p_1), p_2' \in I(p_2). \doms{p_1'}{p_2'} \land \textsc{Dirty}(p_1') = \textsc{Dirty}(p_2')$
\end{theorem}

@@

\begin{case}
$\textsc{Dirty}(p_1) \not= \textsc{Dirty}(p_2)$, in which case $p_1$ is not removed from the plan cache as the $\doms{p_2}{p_1}$ is not within the same dirty set. This is a contradiction
\end{case}

@@

\subsection{Complexity}
\todo{Stengthen complexity argument, as per Sam's feedback}
Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our planning algorithm still operates in exponential time. 
If we removed our restriction on imputation operators to consider not just minimal versus maximal imputations, and instead allowed any arbitrary subset
of attributes to be imputed at a plan node (including those not operated on by that node), then every single operator in the query plan would have a
number of plans exponential in the number of dirty columns.

>>>>

\section{Algorithm}
In order to correctly plan around missing values, we first extend the relational algebra operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) 
with two new operators (\Cref{sec:operators}). We then define the search space (\Cref{sec:search-space}) of plans to encompass non-trivial operations, such as joins and 
aggregations. The decision to place imputation operators is driven by our cost model (\Cref{sec:cost-model}), which characterizes a plan based on estimates for the 
quality of the imputations performed and the runtime performance. Soundness of query execution is provided by our main design invariant, which guarantees traditional relational
algebra operators must never observe a missing value in any attribute that they operate on directly (\Cref{sec:placement}). Our imputation planning algorithm is
agnostic to the type of imputation used (\Cref{sec:imputation}), and we prove that it is optimal, given our search space (\Cref{sec:optimal}). Finally, we show
that while our algorithm is exponential in the number of joins(\Cref{sec:complexity}), in practice, planning times are not a concern.

@@

\subsection{Cost model}
\label{sec:cost-model}
The cost of a plan $q$ is expressed as a tuple $\langle \textsc{Loss}(q), \textsc{Time}(q) \rangle$.
$\textsc{Loss}(q) \in [0,1]$ is an estimate of the amount of information lost by the imputation procedures used.
$\textsc{Time}(q) \in (0, \infty)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates.

@@

This set $P$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all possible trade-offs of $\textsc{Time}$ and $\textsc{Loss}$ for the current query.

@@

The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ relies on having an accurate cardinality estimate of a query plan $q$ and its sub-plans. 
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
For simplicity, each of the logical nodes in a query plan points to a set of histograms.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for the new operation in the plan.
\Cref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.

For each sub-plan, we keep track of the estimated cardinality by using the histograms associated with
that plan. We use this estimate to compute $\text{Loss}(Q)$, a measure averaged over imputation operations, as follows.

@@

To compute $\textsc{Time}(Q)$, an additive measure of runtime cost, we retain a simple set of heuristics used by a standard 
database.

@@

In general, evaluating the time complexity for different \textit{Impute} operators is
complicated, as the properties of
different algorithms can vary significantly, the underlying I/O cost
is not easily extracted (especially in the case that the entire set of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
not even consider CPU cost explicitly). For example, \cite{martin1995time} find that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). Indeed, the \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree. In
practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
plans).

@@

Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a null value, there are cases where an imputation operator is required. To track these cases, each query plan $q$ is associated with a set of dirty attributes $\textsc{Dirty}(q)$. An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain null. We compute a dirty set for each table using the table histograms, which track the number of null values in each column. The dirty set for a query plan can be computed recursively as follows:

@@

Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
At a high level, a plan cache is a mapping from a set of tables to a set of dirty-set specific dominating plans that perform the required joins and filters over these tables.
We store multiple plans rather than a single plan because each plan has a distinct set of dirty attributes, each of which maintains its own Pareto frontier. 
Details of the plan-cache semantics are shown in~\Cref{fig:semantics-plan-cache}.

\begin{figure}
  \begin{align*}
    \llbracket Q[T] \rrbracket &= \begin{cases}
      P & (T, P) \in Q \\
      \emptyset & \text{otherwise} \\
    \end{cases} \\
    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned} Q \cup \{(T,\ P') \} \end{aligned}\\
                                            \text{ where } \\
                                           P' &= \left\{q ~|~ \begin{aligned}
                                                                     q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\
                                                                     \textsc{Dirty}(q) = \textsc{Dirty}(p),\\
                                                                     \forall q' \in \llbracket Q[T] \rrbracket, \textsc{Dirty}(q')=\textsc{Dirty}(q).  \doms{q} {q'} 
                                                                   \end{aligned}
                                           \right\}
  \end{align*}
  \caption{Semantics of a plan cache $Q$}
  \label{fig:semantics-plan-cache}
\end{figure}

@@

The join optimization is based on the algorithm used in System R\cite{blasgen1981system},
but rather than select the best plan at every point, we use our plan cache, which keeps track of the tables joined and the dirty-set specific Pareto frontiers.
The output of the join optimizer is the set of plans which satisfy all join predicates and lie along their respective frontier.

@@

The flexibility of this non-parametric imputation approach aligns well with the spirit of
query optimization, which aims to reconcile performance with a declarative interface to data
manipulation.

@@

\subsection{Optimality}\label{sec:optimal}
We prove the optimality of the final plan selected by showing that our cost-model and algorithm
construct a complete Pareto frontier.

\begin{definition}
$I: \mathcal{P}(D) \rightarrow \mathcal{P}(\{ \delta, \mu \})$, where
$D$ is the dirty set for a relation, is an abstract imputation operator mapping a set of
attributes with missing values to a set of imputation operations.

@@

\begin{theorem}\label{theorem:no-bad-sit}
By computing dirty-set specific Pareto frontiers, it is never the case that we discard a plan $p_1$ s.t. $\doms{p_2}{p_1}$, but 
$\exists p_1' \in I(p_1), p_2' \in I(p_2). \doms{p_1'}{p_2'} \land \textsc{Dirty}(p_1') = \textsc{Dirty}(p_2') \land p_1' \not \in I(p_2)$
\end{theorem}

@@

\begin{case}
$\textsc{Dirty}(p_1) \not= \textsc{Dirty}(p_2)$, in which case $p_1$ is not removed from the plan cache as the $\doms{p_2}{p_1}$ relation does not occur within the same dirty set. This is a contradiction
\end{case}

@@

\subsection{Complexity}\label{sec:complexity}
\todo{Stengthen complexity argument, as per Sam's feedback}
Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our planning algorithm is still exponential in the number of joins. 
If we removed our restriction on imputation operators to consider not just minimal versus maximal imputations, and instead allowed any arbitrary subset
of attributes to be imputed at a plan node (including those not operated on by that node), then every single operator in the query plan would have a
number of plans exponential in the number of dirty columns.
